---
title: "Summer Olympics Performance Analysis"
author: "Furong Bai"
date: "2017Äê4ÔÂ27ÈÕ"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
library(shiny)
# uncomment following if haven't install "gridExtra" package before
# install.packages("gridExtra")
library(gridExtra)
```

## PROBLEM STATEMENT 

Many people like me enjoy watching the Olympic Games, having the valuable and fun time to watch the world top athletes to compete in various sports. People are proud of the success of their country¡¯s athletes, thus number of medals won of a nation is important to national prestige. A glance at the by-country medal tables can clearly tell us that not all the nations have the same ability to win the game. While some countries have dominated the medals in specific sports, others are hardly come into the picture. Hence, there is one question we may want to ask while viewing the games: what makes these countries so successful in the Olympic games? 

In this exercise, I explored Summer Olympics data from 1896-2012, in order to understand the performance of leading countries over year. The objective is to find answers to the following questions:

* What are the top 10 countries in the history of Summer Olympics?
* How're their performances change over time?
* Which gender won more medals in summer Olympic? Did the pattern changed over time?
* What is the impact of socioeconomic variables on national performance in Summer Olympic games? (GDP, Population & per capita GDP)

```{r, echo=FALSE}
# input dataset 
# change the file location if needed
summer <- read.csv("summer.csv")
dictionary <- read.csv("dictionary.csv")
```


## Data Preparation 

#### Explore the data sets
First Check the structure of the data.Original data sets were from Kaggle. 
```{r, echo=FALSE, results="hide"}
# understand the structure of the data
head(summer)
str(summer)
```
"summer" data-set listed information for each medal winner and each medal won. It contained 31165 observations and 9 variable as following:

1. Year: year of the Olympics
2. City: host city
3. Sport: type of sport
4. Discipline: field of sport
5. Athlete: athlete's name
6. Country: which country the athlete belonged to
7. Gender: athlete's gender
8. Event: specific Olympic event
9. Medal: type of medal won: gold, silver or bronze


```{r, echo=FALSE, results="hide"}
head(dictionary)
str(dictionary)
```
"dictionary" data-set included the general information of participated countries. It contained 201 observations and 4 variables listed as following:

1. Country: country name
2. Code: country code
3. Population: country's population
4. GDP.per.Capita: per capita GDP in the country

  
#### Missing values & outliers
Then check the missing values and outliers within the data.
```{r}
sapply(summer, function(x) sum(is.na(x)))
sapply(dictionary, function(x) sum(is.na(x)))
```
* no missing values in "summer" data set 
* "dictionary" data set contained missing values: 5 missing in Population and 25 missing in GDP.per.Capita

Boxplots show that outliers or extreme values may exist in the data-set.
```{r echo=FALSE, warning=FALSE,fig.width=10, fig.height=1.5}
# construct boxplot to check outliers for numerical variables
dictionary %>% ggplot(aes(x = 1, y = Population)) + geom_boxplot() + coord_flip() 
dictionary %>% ggplot(aes(x = 1, y = GDP.per.Capita)) + geom_boxplot() + coord_flip()
```

* Following showed the extreme values in Population:
```{r echo=FALSE}
# filter out extreme values in Population
filter(dictionary,Population>1000000000)
```

* Following showed the extreme values in GDP.per.Capita:
```{r echo=FALSE}
# filter out extreme values in GDP.per.Capita?
filter(dictionary,GDP.per.Capita>62500)
```

* Values for these outliers were reasonable. 
    + Two extreme values in Population were India and China. 
    + Four extreme values in per capita GDP were Luxembourg, Norway, Qatar and Switzerland. 
* Their values were actual and make sense 
* Thus I didn't exclude these entries in analysis.



## Data Analysis 

#### What are the top 10 countries in history of Olympics?
```{r analysis, echo=FALSE}
# Add the count variable to make changes to the orginal dataset
count <- rep(1,nrow(summer))
summer <- cbind(summer,count)

# Summarize total medal counts by Country by Medal: by_country_medel
by_country_medal <- summer %>% 
  group_by(Country, Medal) %>% 
  summarize(Count = sum(count)) 

# make a tidy medal tally
medal_table <- by_country_medal %>% 
  spread(key=Medal,value=Count) %>%    # create columns for each type of medal
  mutate(Total=Gold+Silver+Bronze ) %>%   # create a new column i.e. total medal count
  arrange(desc(Total)) %>%                # arrange rows in descending rank of total medal 
  select(Country,Gold,Silver,Bronze,Total)   # reorder the columns

# Filter the top 10 countries in total amount of medals
top_10 <- medal_table[1:10,]
# Print out the result table
top_10

# save top 10 countries code
top_10_names <- as.character(top_10[1:10,]$Country)

# save top 10 countries code
top_5_names <- as.character(top_10[1:5,]$Country)
```

#### What are the top 10 countries in 2012 Olympics?
* Some countries were leading in 2012 Olympics Medal Tally but didn't show up in overall Medal Tally, such as Russia, China, Japan.
* USA won the most medals in 2012 and overall history.
```{r, echo=FALSE, warning=FALSE}
# Summarize total medal counts by Country by Year: by_country_year
by_country_year <-  summer %>% 
  group_by(Country, Year) %>%
  summarize(TotalMedal = sum(count))

# Filter the 2012 Olympics and arrage in medal rank
by_country_year_2012 <-  by_country_year %>% 
  filter(Year == 2012) %>%
  arrange(desc(TotalMedal))

# Filter the top 10 nations in 2012 Olympics  
top10_2012 <- by_country_year_2012[1:10,]  
top10_2012
```

```{r echo=FALSE, results="hide"}
by_country_year_medal <-  summer %>% 
  group_by(Country, Year, Medal) %>%
  summarize(Count = sum(count)) 
```

## Visualization

#### Which countries won the most medals in history of Olympics?
```{r, echo=FALSE, fig.width=10, fig.height=7}
# filter out top 10 countries from by_country_medal table
by_country_medal_top10 <- by_country_medal %>% filter(Country %in% top_10_names) 

# plot top 10 countries' medal count
by_country_medal_top10 %>% 
  ggplot(aes(x=reorder(Country,Count),y=Count, fill=Medal)) + 
  geom_bar(stat="identity") +
  ggtitle("Bar Chart of Medal Counts For Top 10 Countries") +
  labs(x="Country",y="Medal Count") +
  theme(plot.title = element_text(face="bold", size=14, hjust=0.5)) +
  theme(axis.title = element_text(face="bold")) 

```

The above graph shows the leading countries in medal tally. Followings are some findings in the graph:

* U.S. won the most amount of medals in history of summer Olympics. Its total medals was above 4500 and was about double of 2nd country in medal tally. Gold medals occupied the highest share compared to silver and bronze.
* Although Soviet Union broke down in 1980s, it still won 2nd highest amount of medals in the history of Olympics. 



#### What was the top 10 countries in each year?

* In different year, top 10 countries in the medal tally changed.
* Some countries dropped out of the top list over time and some countries emerged into top list in recent years.
* **Year 1916 1940, & 1944 didn't hold Olympics due to the world war.**
* Select year you want to check which are the top 10 countries in that year.
```{r, echo = FALSE, fig.width=10, fig.height=7}
sliderInput(inputId = "year", label = "Select Year:", min = 1896, max = 2012, value = 2012, step=4, width="200%")

renderPlot({
  top10_selected_year <- by_country_year %>% 
    filter(Year == input$year) %>% 
    arrange(desc(TotalMedal))
  top10_selected_year <- top10_selected_year[1:10,]
  
  appdata <- by_country_year_medal %>% filter(Country %in% as.character(top10_selected_year$Country))

  # draw the bar-chart for year slected
  ggplot(appdata,aes(x=reorder(Country, Count), y=Count, fill=Medal)) +
    geom_bar(stat="identity")+
    ggtitle(paste("Bar Chart of Medal Counts For Top 10 Countries in Year ", input$year)) +
  labs(x="Country",y="Medal Count") +
  theme(plot.title = element_text(face="bold", size=14, hjust=0.5)) +
  theme(axis.title = element_text(face="bold")) 
})
```

#### How did performance of top 5 countries in Olympic history change over time?
```{r perf_change, echo=FALSE,fig.width=10, fig.height=5}

# Filter the top 10 countries in total amount of medals  
by_country_year_top5 <-  by_country_year %>% filter(Country %in% top_5_names)

# plot scatter graph with line to show tendency in change
by_country_year_top5 %>% 
  ggplot(aes(x=Year,y=TotalMedal,group=Country,col=Country)) + 
  geom_point() + 
  geom_line(size=1) +
  labs(x="Year",y="Total Medal Count") +
  ggtitle("Change of Top 5 Countries' Total Medal Count Over Time") +
  theme(plot.title = element_text(face="bold", size=14, hjust=0.5)) +
  theme(axis.title = element_text(face="bold")) 

```

As we can see from the graph above:

* Most of these countries attended Olympic games since invention of the game. 
* USA was keeping consistently high performer over time.
* Soviet Union won about 450 medals in 1980, highest ever among all summer Olympic games.
* China didn't come into the picture maybe because it joined Olympics late.


#### What makes some countries so succefful in Olymics?

* Is there an relationship between some social-economic variables and total medal won?
* Focus on Year 2012.
* Assumption:
    + With larger population, the country may have larger pool of talents.
    + With greater GDP per capita, the country may have greater ability to invest on sports.
  
* Graphs show the scatter plot and relationship between two social-economic variables and total medals.
```{r relationship, echo=FALSE, warning=FALSE, message=FALSE,fig.width=10, fig.height=7}
# Adjust the variable name in "dictionary" corresponding to "summer" because we want to joint two tables later
names(dictionary)[1] <- "CountryFullName"
names(dictionary)[2] <- "Country"

# Make table of Countries' TotalMedal and socialeconomic variables
country_table <- by_country_year %>% 
  inner_join(dictionary,by=c("Country")) %>% # joint two tables based on Country Code
  filter(Year==2012) %>% # Filter out data in Year 2012
  mutate(Pop.million=Population/1000000 ) 

# Relationship between Popoulation & TotalMedal 
p1 <-ggplot(country_table, mapping=aes(x=Pop.million, y = TotalMedal)) +
    geom_point(na.rm=T) +  
    scale_x_log10(expression("Population in million people (log10)")) +
    geom_smooth(se=F,na.rm=T) +
    stat_smooth(method="lm", col="red", se=F, size=0.5,na.rm=T) +
    labs(y="Total Medal Count") +
    ggtitle("Relationship Between Population & Medal Count") +
    theme(plot.title = element_text(face="bold", size=12, hjust=0.5)) +
    theme(axis.title = element_text(face="bold")) 

# Relationship between GDP per capita & TotalMedal 
p2 <-ggplot(country_table, mapping=aes(x=GDP.per.Capita, y = TotalMedal)) +
    geom_point(na.rm=T) +  
    scale_x_log10(expression("GDP.per.Capita (log10)")) +
    geom_smooth(se=F,na.rm=T) +
    stat_smooth(method="lm", col="red", se=F, size=0.5,na.rm=T) +
    labs(y="Total Medal Count") +
    ggtitle("Relationship Between per capita GDP & Medal Count") +
    theme(plot.title = element_text(face="bold", size=12, hjust=0.5)) +
    theme(axis.title = element_text(face="bold")) 

grid.arrange(p1, p2, ncol=2)
```

* Finally, tit a linear regression model, the output of the model is listed below.
    + Intercepts for log(Pop.million) - 14.563;
    + Intercepts for log(Pop.million) - 16.551;
    + Detailed results were listed below:
```{r, echo=FALSE}
#Fitting a linear regression model 
model <- lm(TotalMedal ~ log(Pop.million) + log(GDP.per.Capita), data=country_table)
summary(model)
```


